TacticToe: Learning the Game of Tactics

### 2018-02-14 tactictoe_jv.tex
case study: wip

\section{Case Study}\todo{comment on the case study}
In this section, we compare \tactictoe proof script with human proof script.

Goal: \texttt{IS\_GCD\_UNIQUE} in theory \texttt{pred\_set}).
\begin{lstlisting}[language=SMLSmall]
!a b c d. is_gcd a b c wedge is_gcd a b d ==> (c = d)
\end{lstlisting}

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall]
PROVE_TAC[IS_GCD, DIVIDES_ANTISYM]
\end{lstlisting}

\vspace{5mm}

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall]
STRIP_TAC THEN
REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN
REPEAT Cases THENL
  [METIS_TAC [],
   REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, compute_divides] THEN
     METIS_TAC [NOT_SUC],
   METIS_TAC [NOT_SUC, DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0],
   METIS_TAC [],
   RW_TAC numLib.arith_ss [divides_def],
   METIS_TAC [DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0]]
\end{lstlisting}

@@

Goal: \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set})
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

Original \holfour proof script (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}

\vspace{5mm}

Make it match the new version.
\tactictoe proof script (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

Goal
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}

\vspace{5mm}

Original \holfour proof script (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

%Name: 224 LUPDATE_APPEND2
%Original proof time:

%val LUPDATE_APPEND2 = Q.store_thm("LUPDATE_APPEND2",
%   `!l1 l2 n x.
%      LENGTH l1 <= n ==>
%      (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n-LENGTH l1) l2))`,
%);
\vspace{5mm}
Make it match the new version.
\tactictoe proof script (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
fs is a \sml binding for FULL\_SIMP\_TAC (srw\_ss ()) and rw is short for ??.

If a discovered proof is short, fast, maintainable and
understandable, it could replace its original in the \holfour standard library.

\todoi{Side effect:
+ could be also used to simplify existing proof scripts. }

Interactive usage example.

\begin{lstlisting}[language=SMLSmall]
tactictoe ([], ``countable (UNIV:num set)``);

let fun RWTAC thms = SRW_TAC [] thms in
  RWTAC [fetch "pred_set""countable_def", fetch "pred_set" "INJ_DEF"] THEN
  Q.EXISTS_TAC `\\x.x` THEN
  METIS_TAC []
end
\end{lstlisting}

\begin{lstlisting}[language=SMLSmall]
tactictoe ([], ``count (n+m) DIFF count n = IMAGE ($+n) (count m)``);

SRW_TAC [ARITH_ss] [fetch "pred_set" "EXTENSION", EQ_IMP_THM] THEN
Q.EXISTS_TAC `x - n` THEN
SRW_TAC [ARITH_ss] []
\end{lstlisting}

>>>>

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be part of a \holfour 
development, it interesting to compare their quality with the original human 
proofs. The quality of a proof in \holfour can be measured in terms of length, 
speed, readability and maintainability. We study these properties in three 
examples taken from our full-scale experiment.

In Example~\ref{ex:cs1},

\begin{example}\label{ex:cs1} \texttt{IS\_GCD\_UNIQUE} in theory \texttt{gcd}
\begin{lstlisting}[language=SMLSmall]
!a b c d. is_gcd a b c wedge is_gcd a b d ==> (c = d)
\end{lstlisting}
Human proof (5 milliseconds)
\begin{lstlisting}[language=SMLSmall]
PROVE_TAC[IS_GCD, DIVIDES_ANTISYM]
\end{lstlisting}
\tactictoe proof (80 milliseconds)
\begin{lstlisting}[language=SMLSmall]
STRIP_TAC THEN
REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN
REPEAT Cases THENL
  [METIS_TAC [],
   REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, compute_divides] THEN 
     METIS_TAC [NOT_SUC],
   METIS_TAC [NOT_SUC, DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0],
   METIS_TAC [],
   RW_TAC numLib.arith_ss [divides_def],
   METIS_TAC [DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0]]
\end{lstlisting}
\end{example}

@@

\begin{example} \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set}
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}
Human (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}
\tactictoe (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}
\end{example}

\begin{example} \texttt{LUPDATE\_APPEND2} in theory \texttt{rich\_list}
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}
Human (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}
\tactictoe (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
\end{example}

fs is a \sml binding for FULL\_SIMP\_TAC (srw\_ss ()) and rw is short for ??.
If a discovered proof is short, fast, maintainable and
understandable, it could replace its original in the \holfour standard library.

To conclude, we also demonstrate the capabilities of \tactictoe on two goals 
that do not have a proof in \holfour.

The first goal, in Example~, states that the set of 
natural numbers is countable. \tactictoe starts from the definition of 
countability and injection then instantiate the result with the trivial 
injection and finalize the proof using the \metis. Here,
\tactictoe is able to find an instantiation for an existential quantifier 
because it is a common one. If a more complicated term was needed, \tactictoe 
would try to find an alternative proof that do not rely on an original
instantiation. This weakness could be mitigated by the introduction of term 
prediction in a future version of \tactictoe.

\begin{example}
\begin{lstlisting}[language=SMLSmall]
countable (UNIV:num set)
let fun RWTAC thms = SRW_TAC [] thms in
  RWTAC [countable_def, INJ_DEF] THEN Q.EXISTS_TAC `\\x.x` THEN METIS_TAC []
end
\end{lstlisting}
\end{example}

The second goal, in Example~, \todoi{TG to RK: I am not sure what the count 
constant is doing.}

\begin{example}
\begin{lstlisting}[language=SMLSmall]
count (n+m) DIFF count n = IMAGE ($+n) (count m)
SRW_TAC [ARITH_ss] [fetch "pred_set" "EXTENSION", EQ_IMP_THM] THEN
Q.EXISTS_TAC `x - n` THEN
SRW_TAC [ARITH_ss] []
\end{lstlisting}
\end{example}
